---
title: "Gait cycle selection in treadmill running data: Does it matter?"
author:
  - name: Aaron S. Fox
    affiliation: Centre for Sport Research
    footnote: 1
  - name: Meghan F. Keast
    affiliation: Centre for Sport Research
  - name: John Warmenhoven
    affiliation: UNSW
  - name: Jason Bonacci
    affiliation: Centre for Sport Research
address:
  - code: Centre for Sport Research
    address: Centre for Sport Research, School of Exercise and Nutrition Sciences, Deakin University, Geelong, Australia
  - code: UNSW
    address: School of Engineering and Information Technology, University of New South Wales, Canberra, Australia
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Insert abstract...

journal: ar$\chi$iv
date: "Version 1.0:`r Sys.Date()`"
bibliography: mybibfile.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

#Multiple knit options - in YAML
# knit: (function(inputFile, encoding) {
#   rmarkdown::render(inputFile, encoding = encoding,
#   output_format = "all") })

```
                                                  
# Introduction


- Variety of methods for selecting gait cycles from treadmill-based studies of running...


- Common process within experimental studies is to collect a certain number of gait cycles and use the average of this biomechanical data to compute a representative mean for the individual...


- There is a need to understand the effect of gait cycle number on biomechanical outcome variables from a time cost perspective. We can somewhat easily collect a significant number of gait cycles within a laboratory environment in a treadmill-based test. Having participants settle into a steady state through continuous running over X time period is in fact recommended (?). However, there are potential time savings to be made within analysis protocols should we need to process fewer gait cycles (i.e. time within IK-like processes, time with data cleaning etc.). On the flip side, should only a small number of gait cycles result in 'unstable' biomechanical variables that aren't necessarily representative of an individuals running biomechanics, this would suggest a need to collect and process more gait cycles in calculating a representative mean for the individual... 



Oliveira and Pirscoveanu[@Oliveira2021] recently examined the typical number of gait cycles using in running biomechanics studies. On average, studies used 12 cycles per runner to describe running biomechanics, while Very few (5 out of 56 studies) used more than 10 cycles [@Oliveira2021]. 


Oliveira and Pirscoveanu[@Oliveira2021] subsequently performed a study investigating the impact of sample size and the number of gait cycles used on biomechanical outcomes

***TODO: outline major findings*** 

The findings from their study are, however, specific to overground running and the specific set of variables analysed (i.e. foot contact time, loading rate, peak vertical force, peak braking force, running speed and foot contact angle). Treadmill-based studies of running are common across the biomechanical literature ***ADD REFS***. It is plausible that treadmill running may incur a different pattern of 'stability' with respect to the number of gait cycles needed for analyses, resulting in variable recommendations to those presented by Oliveira and Pirscoveanu[@Oliveira2021]. Further, Oliveira and Pirscoveanu[@Oliveira2021] did not examine lower limb kinematic variables commonly used in gait biomechanics studies. These kinematic variables are presented as both 'zero-dimensional' (0D; e.g. peak values) and 'one-dimensional' (1D; e.g. time-normalised kinematic waveform) variables across biomechanical studies ***ADD REFS***. Additional analyses of these common kinematic variables in both their 0D and 1D forms may, again, yield additional information with respect to the number of gait cycles required in biomechanical research. 


We sought to understand the impact of the number of gait cycles selected, along with where these are selected from during a continuous bout of treadmill running, on biomechanical data and ... how this may influence conclusions...

First, we used sequential analysis to understand the stability of biomechanical variables as gait cycle number is increased.

Second, we examined how representative varying samples of gait cycles were to an individuals entire bout of continuous treadmill running.

Third, we examined whether sampling gait cycles of varying number from different sections of an individuals bout of continuous treadmill running could influence biomechanical outcome variables.

Fourth, we examined whether varying samples of gait cycles could influence our capacity to detect biomechanical differences between conditions — specifically, between running at two different speeds.


# Methods


We used the public dataset of treadmill running biomechanics provided by Fukuchi et al.[@Fukuchi2017]. The specific details of this dataset can be found in the associated paper [@Fukuchi2017]. Briefly, this dataset contains lower-extremity kinematics and kinetics of 28 regular runners (27 male, 1 female; age = 34.8 ± 6.7 years; height = 176.0 ± 6.8 cm; mass = 69.6 ± 7.7 kg; running experience = 8.5 ± 7.0 years; running pace = 4.1 ± 0.4 min/km) [@Fukuchi2017]. Running kinematics were collected using a 12-camera 3D motion capture system (Raptor-4, Motion Analysis, Santa Rosa, CA, United States) and ground reaction force (GRF) data via an instrumented dual-belt treamill (FIT, Bertec, Columbus, OH, United States) [@Fukuchi2017]. Participants ran on the treadmill at three designated speeds (2.5m·s^-1^, 3.5m·s^-1^ and 4.5m·s^-1^), during which a three-minute accommodation period was provided followed by a 30-second data collection period [@Fukuchi2017]. 


- Within our work, we only used data collected from the 2.5m·s^-1^ and 4.5m·s^-1^ trials. [maybe? - probably not...]

We processed the experimental data from Fukuchi et al.[@Fukuchi2017] using OpenSim 4.0 [@Delp2007]. Segment geometry of the generic musculoskeletal model of the pelvis and lower limb provided by Lai et al.[@Lai2017] were scaled for each participant using their static calibration trial, which was also used as a reference for adjusting marker positions on the model. Lower limb joint angles were calculated using filtered (10Hz low-pass 4^th^ order Butterworth) marker trajectory data within inverse kinematics analysis. GRF data were also filtered using the same cut-off frequency and filter. The filtering procedures reflected those originally performed by Fukuchi et al.[@Fukuchi2017]. Foot strike and toe-off events were determined when the vertical GRF crossed a 20N threshold, in line with the original work on this dataset [@Fukuchi2017].



- 0D and 1D variables...outline how these are calculated and what they are...
- Data from right limb


We used sequential analysis to determine the minimum number of gait cycles required to reach a 'stable' representative mean within the number of gait cycles extracted from the 30-second period of treadmill running. Sequential analysis was individually applied across the different running speeds and kinematic variables (both 0D and 1D variants) examined. We used multiple bandwidths to determine varying degrees of 'stability'. This included ± 0.25 of one standard deviation from the target mean, in line with existing studies applying sequential analysis to biomechanical data ***ADD REFS***; but expanded this to ± 0.50, 0.75 and 1.00 of one standard deviation from the target mean. This additional analysis can provide various thresholds for which researchers can choose to apply to their own studies. For 0D variables, the sequential discrete data points across all gait cycles from each participant were averaged to create the target stability, from which the various standard deviation bandwidths were applied to. To be considered 'stable,' a 0D variable had to fall and remain within the bandwidth for the rest of the gait cycles. For 1D variables, the same technique was applied to each time-normalised point along the continuous kinematic waveform. To be considered 'stable,' a all points along the 1D kinematic waveform had to fall and remain within the bandwidth for the rest of the gait cycles. The number of gait cycles where stability was reached across the various running speeds, kinematic variables and stability thresholds were extracted and descriptive statistics (i.e. mean, median, inter-quartile ranges, total range) were calculated. The proportion of participants reaching stability at each gait cycle number was also calculated.


# Results





``` {r cars}

#Test some r code
summary(cars)

```




# Discussion

...

Our fourth question perhaps presents the most practically meaningful analysis of this work, in understanding how the number and selection of gait cycles could impact the conclusions of a comparative study using a paired design...

